### 2012-11-22 paper/isca40format/DistributedMsi.tex
before reorging scheduler etc

In order to transfer the Global MSI protocol into a distribute MSI protocol, we
need to make sure that a method can only read and write the state of the cache
with which the method is associated, and it affects the state of other caches
only by sending messages. In order to read the state of a child's cache we
maintain a shadow state of in the form of a directory. Thus the coherence state
of a cache $c$, in addition to the $c.state[a]$ field, also contains a
directory field containing the shadow state of each child's state field, \ie,
$c.dir[c'][a]$, for child $c'$.

We extend our sequential language for writing methods with two communication
primitives: \send{} $(Req|Resp|Data)\langle c_1 \rightarrow c_2, a,
(x|data)\rangle$ and \receive{} $(Req|Resp|Data)(\langle c_1 \rightarrow c_2,
a, (x|data)\rangle$ where caches $c_1$ and $c_2$ must satisfy the parent-child
relation. Both of these primitives are \emph{blocking}, that is, \send{} may
block because the network is clogged, \ie, has no buffer space to hold a
message and a \receive{} may block because the message has not arrived or cache
has no resources to process it. These primitives make our sequential language
methods ``suspensive'. Request or response messages have very different
characteristics and a data message is conceptually always associated with some
response message. We think of each \textsc{coreReq} as launching a thread in
L1. This thread in turn may get processed, send request or response messages,
may get suspended and eventually die. Thus potentially every request message
can launch a new thread in the cache that receives it. A response message,
however, can only die or generate more response messages; it can never generate
a request message. However a response message can also block waiting for the
corresponding data message. Each thread has an address associated with it which
corresponds to the message that created it.

In order to write the distributed MSI protocol using suspensive threads, we will
maintain the following sets of invariants. In the next section, we will show in
the requirements in the network and the bounds on the number of virtual channels
to avoid head of line blocking. We will also show the other resource
requirements, as well as the properties of the thread scheduler that will ensure
these invariants there.

\subsection{Relation between directory and state of the children's caches}
\begin{theorem}
Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
p.dir[c][a] \ge c.state[a]$.  \label{conservative}
\end{theorem}

This ensures that if a cache makes decision about compatibility of states of
its children using the local directory information, then it does not violate the
compatibility of the children's real states.

\subsection{Network properties}
Regarding the \send{} and \receive{} operations, \ie the network, the following
properties can assume the following:
\begin{theorem}
It a thread has to send a message, it will eventually be able to send the
message.
\label{canSend}
\end{theorem}

\begin{theorem}
If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
get an upgrade-to-$y$ response from the parent where $y \ge x$.
\label{willRecv}
\end{theorem}

\begin{theorem}
If a cache sends an downgrade-to-$x$ request to one of its children, it will
eventually get a downgrade-to-$y$ response from the child where $y \le x$.
\label{willRecv2}
\end{theorem}

\subsection{Eviction}
If handling of a request requires eviction of another cache line, then too the
thread handling the request may get suspended. This would happen if all the
lines that can be allocated for this request may accessed by other suspended
threads. But we still assume the following invariant.

\begin{theorem}
If a thread is waiting for a line to be allocated, \ie for any address to become
available for eviction, then some address will eventually become available for
eviction
\label{evictDead}
\end{theorem}

It this is not true, a thread that is waiting for being allocated a free line
will never proceed leading to a deadlock.

\subsection{Data transfer}
Regarding the responses received, the designer can assume the following 2
invariants. These invariants are used to wait for receiving data messages
appropriately, as only some state transfers also cause data transfers.
We will show the exact need for these invariants later.

\begin{theorem}
If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
just before $p$ sends the response is the same as $c.state[a]$ just before $c$
receives the response.
\label{pcSame}
\end{theorem}

\begin{theorem}
If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
receives the response.
\label{cpSame}
\end{theorem}

\subsection{Need for invariants}
The need for these network invariants is intuitive. If, for instance, a thread wants to
send a message in the output channel of the node and is not able to send one
forever, then the thread can not proceed further. If a thread is waiting for a
response (for a request previously sent), and it never gets the response, the
thread can never proceed further. These scenarios create a deadlock.

We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
\ref{conservative} using an example of a 2-level system containing two L1
caches $c_1$ and $c_2$ with a shared L2 cache $p$.

Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the
data transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
sends an upgrade-to-$M$ to $p$. $p$ receives the request, but assumes that
$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant
\ref{pcSame} is violated. So, it does not transfer the data to $c_1$, resulting
in $c_1$ waiting for data, and hence deadlock. A dual condition can be
constructed similarly, in which $c_1$ downgrades its state for address $a$ from
$S$ to $I$, and the parent assumes that $c_1$ is in state $M$, and keeps
waiting for data.

Invariant \ref{conservative} is important because a parent node consults its
directory to decide the children it needs to send a downgrade request to.  Let
$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
an upgrade-to-$M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
is also $I$ and hence send an upgrade-to-$M$ response to $c_1$. But since the real
state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
\ref{singleWriter}.

Thus, if all the invariants above (\ref{conservative}, to \ref{cpSame}) hold,
then the threads given in Figure \ref{realistic} can not deadlock because it
will always be able to a) send a request, b) send back a response, c) get back
a response for a request it sent, d) receive data when required. Though the
threads executing in a cache are actually suspensive, these invariants give an
illusion of the threads being non-suspensive barring the following exceptions.
Here $this$ represents the cache under under consideration.

\begin{enumerate}
\item $this.dir[c][a]$ can keep decreasing while a thread corresponding to
address $a$ is executing.
\item $this.dir[c][a']$ can keep decreasing while a thread which is evicting
line $a'$ is executing.
\item If a thread receives responses from the parent, then $this.state[a]$ can
decrease between the operation of sending a request to the parent and the
operation of receiving a response from the parent.
\end{enumerate}

As can be seen, these interactions are minimal, and can be ignored since they
affect neither the handling of upgrade nor downgrade requests. Hence, the
correctness of the protocol can be argued in terms of a single threaded,
non-suspensive, execution model. The formal proof for this, as well as the
invariants stated below is beyond the scope of this paper.

In order to design the distributed MSI protocol, the following considerations
must be kept in mind.
\begin{enumerate}
\item There are 3 kinds of incoming messages to be handled: request from that parent,
requests from children and responses from children.
\item A downgrade-to-$x$ message might be received from the parent even if the
state of the cache is $ \le x$. This has to be handled.
\item A parent sends a response to the child only when a request is received. But a child can send an voluntary/unsolicited response to the parent during evictions.
\item A request should be sent to the parent only if the current cache's state
is lower than what's desired to handle its incoming request. Similarly, a
request should sent to its child only if the directory's notion of that cache's
state is higher than what's desired.
\item A second request is not sent for the same address from the same cache to the same
destination, until a response is received for the first request.
\item A cache can not evict a line if it is currently waiting for a response from the parent for the address corresponding to the line.
\end{enumerate}

%We always reserve resources, \eg, cache line slot for an address before sending
%an upgrade request for that address.

The handling of messages follow the exact same algorithm as given for the
global MSI protocol in Figure \ref{msi-template} with the caveat that instead
of reading a child's state, the cache $this$ can only read its own directory.
So, it suffices to simply replace $c.state[a]$ by $this.dir[c][a]$ every time
the state of child $c$ is read by $this$.

As we mentioned above, a new handler has to be written for handling voluntary
responses. This is shown in Figure \ref{msi-unsolicited}.

@@

\begin{algorithmic}
\State // Upgrading $this$ by sending request to parent and getting a response
\Proc {\uReqL}{$p, a, x$}
  \If {$this$ is LLC \&\& $this.state[a] = I$}
    \State \send{} \Req{this}{Memory}{a}{M};
    \State \receive{} \Data{Memory}{this}{a}{d};
    \State $this.data[a] \gets d$;
    \State $this.state[a] \gets M$;
  \ElsIf
  \State \send{} \Req{this}{p}{a}{x};
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndIf
\EndProc
\State // Downgrading $this$' child by sending request to child and getting a
response
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\State // $this$ finally upgrading the directory state of its child and sending
a response to its child
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\State // $this$ finally downgrading its state and sending a response to its
parent
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\caption{Distributed MSI protocol. In all the methods $p$ is the parent of
$this$, $c$ is a child of $this$, and the methods work on address $a$ to change
its state (or directory) to $x$. This function only shows the auxiliary
functions and the handler for voluntary responses from children. The handler
for requests are exactly the same as in Figure \ref{msi-template}, with every
occurence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
\label{realistic}
\end{figure}

>>>>

We will take the protocol designer's perspective, and first describe what the
designer has to specify to create a distributed protocol.

Each cache can only read and write its local state. A cache maintains a shadow
 of all its children's states in the form of a directory. The designer can
always assume that the state of an actual child is what is given in the
directory. The directory entry of address $a$ for $this$' child $c$ is denoted
using $this.dir[c][a]$.

The designer has to specify the handlers for \uReq{} and \dReq{}. In the
following, we assume that the cache whose handler the designer is writing is
$this$ and the relevant address is $a$.

In order to read a child's state, the designer simply reads the state of the
directory for that child. But to modify the state of child $c$ to $x$, the
designer specifies a command \send{} \Req{this}{c}{a}{x}, followed by \receive{}
\Resp{c}{this}{a}{y}. Command \send{} \Req{this}{c}{a}{x} can be called only when $x
< this.dir[c][a]$. $y$ denotes a free variable which gets filled with the value of
the state $c$ has changed to. $y$ is guaranteed to be $\le x$. $this.dir[c][a]$
should be updated with this value.

In order to read the data of another node $n$ (parent or child),
the designer issues the command \receive{} \Data{this}{n}{a}{d}, where $d$ is a free
variable that gets filled with the data. To write the data into address $a$ of
another node $n$ (parent or child), the designer issues the command \send{}
\Data{this}{n}{a}{this.data[a]}.

In order to upgrade the state of $this$ to $x$, the designer has to
specify the command \send{} \Req{this}{p}{a}{x} followed by \receive{}
\Resp{p}{this}{a}{y}. Command \send{} \Req{this}{p}{a}{x} can be called only when $x
> this.state[x]$. $y$ denotes a free variable which gets filled with the value of
the state $this$ should be upgraded to. $y$ is guaranteed to be $\ge x$.

Using these commands, the designer can now easily specify a distributed MSI
protocol, by specifying the methods \uReq{} and \dReq{}. Three other
obligations to take care of:
\begin{enumerate}
\item Handler \uReq{}($c, a, x$) can be called even if $x \ge
this.dir[c]$. In this case, it should be dropped.
\item Handler \dReq{}($p, a, x$) can be called even if $x \le
this.state[a]$. In this case, it should be dropped.
\item A new handler, \dRespL{}($c, a, x$), where $c$ is a child of cache $this$,
should be provided, and this informs $this$ that $c$ has changed its state to
$x$.
\end{enumerate}

Using the above set of rules, the designer can specify the distributed MSI protocol as shown in
Figure \ref{realistic}. Handlers \uReq{} and \dReq{} are exactly the same as
shown in Figure \ref{msi-template}, except that whenever the state of a child
$c'$ is read, it should be replaced syntactically by $this.dir[c'][a]$.

@@

\begin{algorithmic}
\State // Upgrading $this$ by sending request to parent and getting a response
\Proc {\uReqL}{$p, a, x$}
  \If {$this$ is LLC \&\& $this.state[a] = I$}
    \State \send{} \Req{this}{Memory}{a}{M};
    \State \receive{} \Data{Memory}{this}{a}{d};
    \State $this.data[a] \gets d$;
    \State $this.state[a] \gets M$;
  \Else
  \State \send{} \Req{this}{p}{a}{x};
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndIf
\EndProc
\State // Downgrading $this$' child by sending request to child and getting a
response
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\State // $this$ finally upgrading the directory state of its child and sending
a response to its child
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\State // $this$ finally downgrading its state and sending a response to its
parent
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\caption{Distributed MSI protocol. In all the methods $p$ is the parent of
$this$, $c$ is a child of $this$, and the methods work on address $a$ to change
its state (or directory) to $x$. This function only shows the auxiliary
functions and the handler for voluntary responses from children. The handler
for requests are exactly the same as in Figure \ref{msi-template}, with every
occurrence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
\label{realistic}
\end{figure}

Even though the designer only specifies the distributed protocol in the manner we have
written in Figure \ref{realistic}, we automatically
create a correct implementation of the cache controller from this
specification.  The commands \send{} and \receive{} get translated into
appropriate request and response messages which get sent to or received from the respective
nodes, and they invoke the corresponding message handlers. Note that in the final
implementation, sending of a message blocks if there is no space in the output
channel and receiving of a message blocks if there is no message available for
reception, though the designer is oblivious to this fact. We will describe how
the specification is translated to an actual implementation, and show why the
implementation is correct.

\subsection{Caches as a system executing suspensive threads}
The behavior of each cache node can be best described as a system of suspensive
threads. These correspond to Miss Status Handling Register (MSHRs) in an actual
implementation. A scheduler creates and schedules threads within the cache.  A
new thread is created whenever a new message arrives in the input channel, and
no other thread is executing. When a thread is executing a \send{} command, then
an actual message is created and sent to the appropriate cache. The thread then
goes into a suspend state till it receives a response message indicating that
the requested state change has taken place. In this case, the suspended thread is
woken up by the scheduler to perform the rest of its actions. Once all of its
actions are complete, a thread dies, and its resources are freed.

Each request goes through several stages of processing in a cache (as shown in
Figure \ref{msi-template}). This is the reason for using the vocabulary of
threads instead of MSHRs -- a thread makes the processing stage a particular
request is currently in, implicit.

Now that we have established the vocabulary of threads, we will proceed to
describe the requirements that a network should guarantee in order to make such
an implementation possible. We will also show the scheduling algorithm that we
generate.

\subsection{Message ordering requirements}
\floatstyle{boxed}
\restylefloat{figure}

%The basic ordering requirement is that responses should not overtake each other
%when they are sent from the same source to the same destination. Similarly,
%requests should not overtake responses from the same source to the same
%destination.

\begin{figure}\small
\begin{requirement}
An incoming response for an address $a$ from a child $c$ should not be
received before another incoming response from the same $c$ and the
same $a$ sent earlier has been received\label{cRespFifo}
\end{requirement}
\begin{requirement}
An incoming request for an address $a$ from a source $n$ should not be
received before another incoming response from the same $n$ and the
same $a$ sent earlier has been received\label{reqNoOvertakeResp}
\end{requirement}
\caption{Ordering requirements}
\label{order}
\end{figure}

Figure \ref{order} gives the exact ordering
requirements between messages transmitted between the caches.

We will illustrate the need for Requirements \ref{cRespFifo} and
\ref{reqNoOvertakeResp} using a 2-level cache hierarchy with two L1 caches $c_1$ and
$c_2$, and one shared L2 cache $p$. Each core can have several outstanding
memory requests.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{complicated}
\caption{Effect of violating Requirement \ref{cRespFifo} (order between responses)}
\label{complicated}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

Let's say $c_1.state[a] = p.dir[c_1][a] = M$, $c_2.state[a] = p.dir[c_2][a] =
I$. $c_2$ gets a load request from its processor for address $a$ and sends an
upgrade request to $p$. $p$ sends a downgrade-to-$S$ request to $c_1$. $c_1$
receives the downgrade-to-$S$ request and downgrades to $S$, sending a downgrade
response to $p$. $c_1$ decides to evict address $a$ to make room for another
address $a'$, because of a request from the core for $a'$ not present in $c_1$.
$c_1$ sends a downgrade-to-$I$ response to $p$. Let's say the second downgrade
to $I$ response reaches $p$ first, violating Requirement \ref{cRespFifo}. $p$
sends an upgrade-to-$S$ response to $c_2$. $c_2$ receives the upgrade response
and changes state to $S$. $p$ then receives the second downgrade-to-$S$ response
from $c_1$, changing its $p.dir[c_1][a]$ to $S$. $c_2$ sends an upgrade-to-$M$
request to $p$ because of a store from its core for $a$. $p$ will send a
downgrade-to-$I$ request to $c_1$, but $c_1$ drops the request as seen in
procedure \dReq{} in Figure \ref{realistic}. So $p$ will never receive a
response from $c_1$ leading to a deadlock. This is illustrated in Figure
\ref{complicated}. In the figure, the timeline moves from left to right.
Whenever the state of the directory or the cache changes, we show the new state.
Messages are shown using arrows from the source to the destination, moving
across time.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{lessComplicated}
\caption{Effect of violating Requirement \ref{reqNoOvertakeResp} (requests overtaking responses)}
\label{lessComplicated}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

Consider another scenario starting with $c_1.state[a] = p.dir[c_1][a] = S$ and
$c_2.state[a] = p.dir[c_2][a] = I$. $c_1$ receives a store request and sends an
upgrade-to-$M$ request to $p$. $p$ sends back an upgrade-to-$M$ response to
$c_1$. $c_2$ gets a load request for address $a$ and sends an upgrade-to-$S$
request to $p$. $p$ sends a request to $S$ to $c_1$. Suppose Requirement
\ref{reqNoOvertakeResp} is violated and the request to $S$ reaches $c_1$ before
the response to $M$. $c_1$ will drop the request since its already in state $S$,
and so $p$ will never get a response leading to a deadlock. This is illustrated
in Figure \ref{lessComplicated}.

These scenarios show that these ordering requirements are indeed necessary for a
correct cache coherence protocol.

\subsection{Resource requirements and blocking restrictions}

\begin{figure}\small
\begin{requirement}
At a cache node, an incoming request from the parent should not be blocked by any
incoming request from its children.\label{cReqNoBlockPReq}
\end{requirement}
\begin{requirement}
At a cache node, an incoming response from any of its children should not
be blocked by any incoming requests.\label{reqNoBlockResp}
\end{requirement}
\begin{requirement}
At a cache node, an incoming response for address $a$ from its parent should not be blocked by
an incoming request from its child for address $a$. \label{reqBlockResp2}
\end{requirement}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

Figure \ref{blocking} shows the restrictions that the input channels should
obey with respect to an incoming message blocking other incoming messages.

\begin{figure}\small
\begin{requirement}
Each cache node has at least one dedicated thread resource for handling only messages
(requests or responses) from the parent\label{dedicate1}
\end{requirement}
\begin{requirement}
Each cache node has at least one dedicated thread resource for handling only responses
(from children or from the parent)\label{dedicate2}
\end{requirement}
\begin{requirement}
Each cache node has at least three thread resources\label{dedicate3}
\end{requirement}
\caption{Dedicated resources in a cache node}
\label{dedicated}
\end{figure}

Figure \ref{dedicated} shows the minimum number of thread resources that should
be present in each cache node. Lack of required thread resouces leads to the same
kind of deadlock scenarios as incorrect blocking does. This is illustrated in the
following:

Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c1$.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{unknown}
\caption{Effect of violating Requirement \ref{cReqNoBlockPReq} (requests blocking responses from parent)}
\label{unknown}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

Let's say Requirement \ref{cReqNoBlockPReq} is not held, and instead some request
from the core is blocking $c_1$ from receiving the request. $c_1$ will never
respond back to $p$. But $p$ will be waiting in procedure \dReqL{}, and will
never start handling the request from $c_1$. This will create a deadlock.  The
same deadlock is also created when $c_1$ is out of thread resources to create a
thread to handle the downgrade request from $p$ (violating Requirement
\ref{dedicate1}). Figure \ref{unknown} depicts this scenario.

Let's extend this scenario further. Instead of violating Requirements
\ref{cReqNoBlockPReq} or \ref{dedicate1}, let $c_1$
send a downgrade-to-$I$ response to $p$. Let's instead assume that Requirement
\ref{reqNoBlockResp} is violated in $p$. So, the downgrade-to-$I$ response from
$c_1$ is not received by $p$, leading to the same deadlock. This is the same if
Requirement \ref{dedicate2} is violated and $p$ is out of thread resources to
create a thread to handle the downgrade response.

The need for Requirement \ref{dedicate3} can be shown easily. Let's say that a
cache has exactly 2 thread resources, one for handling messages from parents,
and one for handling responses from either the parent or the children. Now, if
it receives a request from one of its children, it can not use either of the
thread resources. Thus, at least one more thread resource is needed for
handling requests from children. This thread resource can also be used for
handling other types of incoming message.

\subsection{Scheduler requirements}

We will now list the requirements that a scheduler must obey to give the
illusion of non-suspensive threads and avoid deadlocks.

\begin{figure}\small
\begin{requirement}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{requirement}
\begin{requirement}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from $p$, or
\item a thread in $c$ is handling a request for address
$a$ from one of $c$'s children $c'$, but that thread is not waiting for a
response from $p$
\end{enumerate}
\label{cHandleReq}
\end{requirement}
\caption{Requirements for receiving a request (used by the scheduler)}\label{recvReq}
\label{}
\end{figure}

\begin{figure}\small
\begin{requirement}
Address $a$ present in cache node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}\label{evict}
\caption{Requirements for suspending a thread during choosing an address for eviction}
\end{requirement}
\end{figure}

Requirements \ref{pHandleReq}, \ref{cHandleReq} and \ref{evict} are needed
because once a thread has finished up-to a point in its execution, it can not
re-execute that portion. For instance, consider a thread handling an upgrade
request from a child. Let's say thread $t$ has progressed to the point where it
is simply waiting to send an upgrade response to the child. Now, if the line
gets evicted because it was chosen for replacement, or it got downgraded by the
parent, then thread $t$ can not re-execute its procedure again. To prevent this,
only one thread is allowed to operate on a particular address, except in the
case of downgrade requests from the parent. In this case, a thread is created to
handle the downgrade request from the parent even if there is another thread
handling a request for the same address from a child, if the latter thread is
just waiting for a response from the parent. This is done to avoid deadlocks
like the ones shown in Figure \ref{unknown}.

\begin{figure}\small
\begin{requirement}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled to execute.\label{starvation}
\end{requirement}
\end{figure}

Local starvation-freedom is required in each cache node. If a cache node never
executes a particular suspended thread, even if it keeps getting ready to
execute, then the thread will not progress, leading to a deadlock.

A scheduler can be implemented as described in Figure \ref{scheduler}, obeying
Requirements \ref{pHandleReq} and \ref{cHandleReq}. It should perform some form of
fair arbiration among its suspended threads to avoid starvation (Requirement
\ref{starvation}).

\newcommand{\lWhile}{\textbf{while}}
\newcommand{\lIf}{\textbf{if}}
\newcommand{\lElsIf}{\textbf{else if}}
\newcommand{\lElse}{\textbf{else}}

\floatstyle{plain}
\restylefloat{figure}

\begin{figure}
\small
\begin{boxedminipage}{\linewidth}
\begin{alltt}
\normalfont
\lWhile{} (\(True\)) \bopen
      \lIf (\Resp{c}{n}{a}{x} is in the input channel,
            where \(n = c.parent\)) \bopen
            \lIf (suspended thread \(t\) is waiting for 
                 a response from \(c\) for address \(a\)) \bopen
                   \resume{} \(t\);
            \bclose \lElse \bopen
                   \receive{} \Resp{c}{n}{a}{x};
                   \start{} \dRespL(\(c, n, a, x\));
            \bclose
      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel,
            where \(p = n.parent\)) \bopen
                   // There must be a suspended thread \(t\)
                   waiting for a response from \(p\) for address \(a\)
                   \resume{} \(t\);
            \bclose
      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input channel,
                     where \(n = c.parent\)) \bopen
             \lIf (Local-Property \ref{pHandleReq} is satisfied) \bopen
                   \receive{} \Req{c}{n}{a}{x};
                   \start{} \uReq(\(c, n, a, x\));
             \bclose
      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel,
                  where \(p = n.parent\)) \bopen
             \lIf (Local-Property \ref{cHandleReq} is satisfied) \bopen
                   \receive{} \Req{p}{n}{a}{x};
                   \start{} \dReq(\(p, n, a, x\));
             \bclose
      \bclose \lElsIf (suspended thread \(t\) is ready to execute) 
             \resume{} \(t\);
\bclose
\end{alltt}
\end{boxedminipage}
\caption{Implementation of a scheduler for cache node $n$}
\label{scheduler}
\end{figure}

### 2012-11-22 paper/isca40format/Implementation.tex
before reorging scheduler etc

\section{Implementation overhead}
\label{sec:implementation}
In this section we show how the requirements presented in Section
\ref{sec:network} can be implemented in a realistic system.

@@

Since responses from a child of a cache node should not be blocked at the cache
node by any incoming requests (Requirement\ref{reqNoBlockResp}), the responses
from a cache node at a lower level (children) to a cache node at a higher level
(parent) should not use the same share use any of the request channels. Since a
response is always removed by the system, just one channel, $RespChannel$, is
sufficient for transmitting responses from any child to its parent, in the
whole system.

@@

\subsubsection{Ensuring the ordering requirements}
There are no ordering requirements between requests sent from the same source
to the same destination according to Figure \ref{order}.  However, requests
should not overtake responses to the same address, because of Requirement
\ref{reqNoOvertakeResp}. Requests and responses from a child to its parent come
on different virtual channels. Hence, maintaining Requirement
\ref{reqNoOvertakeResp} for messages from a child to its parent is difficult,
since it's difficult to guarantee point-to-point FIFO ordering the messages
come on different virtual channels.

Point-to-point FIFO ordering for messages from a child to its parent can be
ensured by adding more information to the messages rather than relying on the
underlying network. Each message for address $a$ from child $c$ to its parent
$p$ will contain an additional tag which specifies the state the child was in
at the time of sending the message. The new message becomes \Req{c}{p}{a}{x
\rightarrow y} or \Resp{c}{p}{a}{x \rightarrow y}, where $x$ specifies the
state of the cache at the time the request was sent. At the parent, a thread
for handling this message will be created, and this message will be removed,
only after $p.dir[c][a]$ becomes less than or equal to $x$, where $x$ is the
state of the child at the time the message was sent. The intuition behind why
this works is as follows (we do not give the formal proof here). Requests can
not be sent from a child when there is a pending response from the parent
. Similarly, a response can not be sent from
a child when there is a pending response from the parent. So, a response from the parent can never be in flight
when a request or a response from the child is in flight, since the
parent only sends responses on requests from the child . So the directory state of the parent at the time of
receiving a message can never be higher than the state of the child just before
sending the message, if the message did not overtake a previous response.

One issue with this for responses is that if only one virtual channel is used
for responses from any child to its parent throughout the system is as follows.
A response sent later for an address from a child to its parent can block a
response for the same address from the same child to its parent sent earlier.
This violates Requirement \ref{cRespFifo}. If two separate virtual channel are
used, one for downgrade-to-$S$ responses ($RespChannel_S$) and one for
downgrade-to$I$ responses ($RespChannel_I$), then using the response sent later
will not block an earlier response.

@@

\subsubsection{MSHRs to implement threads}
The threads we describe essentially represent state stored in Miss Status Handling Registers (MSHRs). To
satisfy the requirements of Figure \ref{dedicated}, there must at least be 3
MSHR registers, in which one can be used only for messages from the parent, one
only for responses (both from children and the parent) and one general purpose
MSHR. We used the terms threads to describe our protocol because we give the
illusion of only one thread executing in the system though the threads are
actually suspensive. Hence the software terminology fits our model quite well.

>>>>

\section{Implementation overhead}
\label{sec:implementation}
In this section we show how the requirements presented in Section \ref{sec:DistributedMsi} can be implemented in a realistic system.

@@

Since responses from a child of a cache node should not be blocked at the cache
node by any incoming requests (Requirement \ref{reqNoBlockResp}), the responses
from a cache node at a lower level (children) to a cache node at a higher level
(parent) should not use any of the request channels. Since a
response is always removed by the system, just one channel, $RespChannel$, is
sufficient for transmitting responses from any child to its parent, in the
whole system.

@@

\subsubsection{Ensuring the ordering requirements:}
There are no ordering requirements between requests sent from the same source
to the same destination according to Figure \ref{order}.  However, requests
should not overtake responses to the same address, because of Requirement
\ref{reqNoOvertakeResp}. Requests and responses from a child to its parent come
on different virtual channels. Hence, maintaining Requirement
\ref{reqNoOvertakeResp} for messages from a child to its parent is difficult,
since it's difficult to guarantee point-to-point FIFO ordering of the messages
coming on different virtual channels.

Point-to-point FIFO ordering for messages from a child to its parent can be
ensured by adding more information to the messages rather than relying on the
underlying network. Each message for address $a$ from child $c$ to its parent
$p$ will contain an additional tag which specifies the state the child was in
at the time of sending the message. The new message becomes \Req{c}{p}{a}{x
\rightarrow y} or \Resp{c}{p}{a}{x \rightarrow y}, where $x$ specifies the
state of the cache at the time the request was sent. At the parent, a thread
for handling this message will be created, and this message will be removed,
only after $p.dir[c][a]$ becomes less than or equal to $x$, where $x$ is the
state of the child at the time the message was sent. The intuition behind why
this works is as follows (we do not give the formal proof here). Requests can
not be sent from a child when there is a pending response from the parent. Similarly, a response can not be sent from
a child when there is a pending response from the parent. So, a response from the parent can never be in flight
when a request or a response from the child is in flight, since the
parent only sends responses on requests from the child . So the directory state of the parent at the time of
receiving a message can never be higher than the state of the child just before
sending the message, if the message did not overtake a previous response.

One issue with this for responses is that if only one virtual channel is used
for responses from any child to its parent throughout the system is as follows.
A response sent later for an address from a child to its parent can block a
response for the same address from the same child to its parent sent earlier.
This violates Requirement \ref{cRespFifo}. If two separate virtual channel are
used, one for downgrade-to-$S$ responses ($RespChannel_S$) and one for
downgrade-to-$I$ responses ($RespChannel_I$), then using the response sent later
will not block an earlier response.

@@

\subsubsection{MSHRs to implement threads:}
The threads we describe essentially represent state stored in Miss Status Handling Registers (MSHRs). To
satisfy the requirements of Figure \ref{dedicated}, there must at least be 3
MSHR registers, in which one can be used only for messages from the parent, one
only for responses (both from children and the parent) and one general purpose
MSHR. We used the term threads to describe our protocol because we give the
illusion of only one thread executing in the system though the threads are
actually suspensive. Hence the software terminology fits our model quite well.
